Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-17 draft_16Jan/draft7.tex
Changed abstract and conclusion

\begin{abstract}
  Collecting measurements from underlying artificial, natural, and physical systems through observables to realise a dynamical model with greater descriptive power is a well-established practice. Physical systems can often be modelled by instance with differential equations derived from the laws of physics, if we knew the laws.  However, we often do not fully understand the many complexities of the more intricate systems we interact with on a daily basis.
  In this project we start with the hypothesis that an discrete-time dynamical systems, particularly those which are chaotic generates the data. The general learning problem is then formulated and a well-known approach to the problem, alongside its practical limitations are considered.
  We then introduce the notion of driven dynamical systems and establish the theoretical existence of dynamical systems possessing certain properties that guarantee them to be topologically conjugate ('dynamically equivalent') to the dynamics of the underlying system from which measurements were taken. The methodology is practically implemented by making use of a recurrent neural network and using deep learning methods to train on the recurrent neural network states to show that we may obtain long-term consistent predictions for simple physical systems such as the double pendulum as well as a few other attractors exhibiting chaotic behaviour. The machine learning algorithms for modelling the double pendulum thus far have not been successful when it exhibits chaos. 
  \end{abstract}

@@

Practically, the application of Takens embedding involves learning a map through some technique, and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks (Citations to echo state networks)) by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (Citations here) as there is often no guarantee that the right function was learnt during training.

This thesis involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter~ref{ch.4}). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

Formally, a function $T: U \to U$, where $U$ is some set is  a \emph{discrete-time dynamical system} and its iterates $\{u,Tu,T^2u,\ldots\}$, where $T^n$ denotes the $n$-fold composition of $T$ with itself, describe the evolution of an initial condition $u\in U$ (Note that we frequently drop the brackets and denote $T(u)$ by $Tu$ so as to simplify notations).

@@

\begin{Example} \rm
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$ is of particular interest to us as it is classically used to illustrate the influence of chaos on a system \textbf{(Cite)}. 
  In the graph below~\ref{fig:log_sdic} we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

  \begin{figure}[ht]
    \includegraphics[scale=0.74]{_logistic_sdic.eps}
        \centering
        \captionof{figure}{Figure generated for Logistic Map $x_{n+1}=rx_n(1-x_n)$ with $r=4$ to exhibit the presence of SDIC. Plotted are the values $x_n$ against time $n$ for the the first 50 step with initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.}
        \label{fig:log_sdic}
      \end{figure}
  \ednote{B: Caption and figure adjusted}
  % For an intuitive sense of the fact that the LM exhibits SDIC, consider again the definition and substitute specific values. Let $\delta=0.05$ and for purposes of illustration consider the value $u=0.4$. The green band in~\ref{fig:log_sdic} shades the region $[u-\delta, u+\delta]$ and we can see that regardless the value $v$  and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$
  % specific point, say $u=0.4$. Draw a green neighbourhood $[0.35,0.45]$ of diameters $\delta=0.02$ around the point. It is easy to see that every neighbourhood around $u=0.4$ will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta=0.05$ . 
\end{Example}

@@

\begin{Definition}\rm  
[\bf {Conjugacy \& Semi-Conjugacy}]\label{Dfn_Conjugate}\rm
   Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$ where T has domain $U$ and S has domain $V$; S is said to be semi-conjugate to T. 
\end{Definition}

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e.\ednote{two i.e.'s}, we examine  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

From here we construct a multidimensional observable using the method of stacking previous observations, i.e., we create delay-coordinate map defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
Takens' theorem, in essence, refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{B: I remember you saying this in one of our sessions, but don't have a source for it.}

@@

One of the first results obtained after defining the USP is that every solution will attract different initial conditions towards the component parts of the solution.
If $g$ has the USP, then any solution to $g$ is also a uniform attractor in nonautonomous dynamical systems literature~\cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this project, and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

@@

\begin{Definition}
  [\bf Causal Mapping]\label{Def_CausMap}
  A continuous, surjective map $H:\overleftarrow{U}\to\overleftarrow{X}_U$ such that \[H\circ\tilde{g}_v=\sigma\circ{H}\] holds for all $v \in U$ where $\tilde{g}_v$ maps $(\ldots, u_{-2}, u_{-1})$ to $(\ldots, u_{-2}, u_{-1}, g(v, u_{-1}))$ is called a causal mapping.
\end{Definition}

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. 
\ednote{B: edited the sentence here}

@@

  It is imperative to take note of a subtlety here: we refer to the \emph{map} $H$ as a (causal) embedding, whereas the driven \emph{system} $g$ (causally) embeds another \emph{system} $(U,T)$. \ednote{B: inserted line}

@@

%To achieve a causal embedding embedding not just a causal mapping, it is undesirable to choose a function $g$ that quenches the temporal structure in $u$ by contracting to such a degree that the ability to recover information from the original system is lost completely.
In the above example(\ref{ex_halfux}), the input's temporal variation could not be related to the reachable set as $X_U$ only consisted of a single element; little, if not no, information is encoded.
To obtain a suitably complex function $g$, it is thus desired that the reachable set of a driven system  be large enough to relate to the input. 
This is true even for embedding  the inverse-limit space of a dynamical system. \ednote{B: Expand or refer?}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.35]{_Nonlinearity_Fig_response.eps}
  \centering
\captionof{figure}{A coordinate of a simulated solution $(x_0,x_1,\ldots,x_{5000})$ of a RNN  plotted in red (with parameter $\alpha=0.99$ (where $g$ has the USP) and blue ($a=1$ and $\alpha=1.05$ where $g$ does not have the USP) while the matrices $A$ and $B$ are randomly generated, and $B$ has unit spectral radius. 
Reproduced from \cite{Manju_Nonlinearity}. }
\label{fig:memloss_conttime} 
\end{figure} 
\ednote{B: How can I explain the black line? I don't think this makes sense to the reader immediately.}

@@

In \cite{manjunath2021universal} it is illustrated in an empirical fashion in that the map $G_T$ describes dynamics which are less functionally complex than that of $T$ or of the map $\Phi_{2d,\theta}$. This is done by implementing a Recurrent Neural Network (RNN), which is discussed in~\ref{ch5}. \ednote{B: Should I add in that we'll be using Pearson coefficient?}

@@

The reasons for taking this roundabout approach remain to be discussed in \ref{subs_LearnGamma}, but a pat answer may immediately be given: When $\Gamma$ has been learnt, the system can be driven autonomously and then $G_T$ is known anyway. We formalise this with a theorem.

@@

Secondly, the function $\Gamma$ is known to be stable. Learning $\Gamma$ makes use of equations \ref{eqns_from_data} which in turn employs the driven system $g$ possessing the USP, and offers distinct advantages with regards to stability in the presence of perturbation. It is desirable that inputs that differ only slightly would have only ‘slight’ effects on the output of the system. Moreover, one would hope to prevent numerical errors originating from input and measurement noise (see~\cite[Th. 5]{manjunath2021universal}). 
As the data fed into the system is a function of both the input given and the parameters of the system itself (i.e. the value a, $\alpha$ in $\tanh$), we exploit both input-related and parameter-related stability. Both these notions are defined by way of the continuity of an encoding map of an input and design parameter respectively in \cite{manjunath2020stability}.\ednote{B: Is this satisfactory?}

@@

Thirdly, learning a chaotic map on a set(here $Y_T$) with even a small measure of noise  means that the learnt system accepts arguments out of the set (Problems similar to Takens, cf.~\ref{sect_takenslimits}). 
We thus have a map $G_T^+$ acting on $Y_T^+$ and since $Y_T$ is not guaranteed to be an attractor of $G_T$ this could lead to errors.
If one learns $G_T$ directly, we are in danger of replicating the problems arising for Takens.  To see this, consider Fig.~\ref{fig:YtGtFailure}.

\begin{figure}[ht]
  \includegraphics[scale=0.25]{_YTerrors.eps}
  \centering
  \captionof{figure}{Diagram of potential errors developing when $G_T$ is learnt.}
  \label{fig:YtGtFailure}
\end{figure}

To circumvent this: if a driven system with a saturation function\ednote{B: What is a saturation function?} is used, we can indirectly (via $g$) ensure the existence of an attractor $A$ of the map which implements $G_T$, and containing $Y_T^+$.  \ednote{B: I'm not sure this sentence makes complete sense.}
By defining the dynamics to $A$, this prevents large errors from occuring.  
More details on the global dissipativity of $g$, as it is termed there, may be found in~\cite{Supp}.

@@

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\ref{eqns_from_data}. To reach this point, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value and lies within the range $[-1,1]$ and required zero-padding \ednote{M: this zero padding is something that we need to tell the examiner}, we continue by driving the system with external input through the function $g$.

@@

We make use of a Recurrent Neural Network (RNN) in our implementation to project the data on to a higher dimensional space, but we do not train the RNN in the conventional manner. \ednote{B: Should we say what the conventional manner is?} A few words about the RNNs: RNNs are widely used in the literature for temporal processing of data as a type of deep learning technique. They take the form of equations considered to be discrete-time version of continuous dynamical systems modelling a neural network of neurons \cite{jaeger2001echo}.

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: Added so that FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: Note! We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

The two model equations for learning from data (\eqref{eqns_from_data}) allows us to determine what the state $x_{n+1}$ at time $n+1$ will be. 
We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.95\linewidth]{_dp_success_4coords_traj.eps} 
  \captionof{figure}{Predicted(red) and actual(blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads. From top to bottom are as follows: $x$-coordinate of topmost pendulum head, $y$-coordinate of topmost pendulum head, $x$-coordinate of lower pendulum head, $y$-coordinate of lower pendulum head. 
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 300 timesteps and stay close up until about 600 seconds.  } 
  \label{fig:dp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{_dp_success_4coords_hist.eps}
  \captionof{figure}{Densities of predicted and actual values of $x$- and $y$-coordinate of pendulum heads.}
  \label{fig:dp_success_density}
 \end{figure}

>>>>

\begin{abstract}
  It is a well-established practice to collect measurements from underlying artificial, natural, and physical systems through observables in order to realise a dynamical model with greater descriptive power. 
  Many data-driven approaches to forecast the future evolution of systems exist. However, many of these methods rely on an initial understanding of the system and we often do not fully understand the many complexities of more intricate systems that we interact with on a daily basis.

  In this project, the base assumption is that data is originates from  some discrete-time dynamical system, particularly one of a chaotic nature but no assumptions are made pertaining to a specific dynamical system.
  A general learning problem is formulated and a well-known approach to the problem, alongside its practical limitations, is then considered.
  The notion of driven dynamical systems is introduced, and we subsequently establish the theoretical existence of a dynamical system possessing certain properties which guarantees it to be topologically conjugate (`dynamically equivalent`) to the dynamics of the underlying system from which measurements were originally taken. 
  Our methodology is implemented practically by making use of recurrent neural networks, a form of deep learning, to learn a map topologically equivalent to the underlying system that forecasts future states.
  In so doing we present the workability of an approach to obtain long-term topological and statistically consistent predictions for simple physical systems as well as a some chaotic attractors.
  \end{abstract}

@@

Practically, the application of Takens embedding involves learning a map through some technique, and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks \cite{jaeger2001echo, jaeger2004harnessing} by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (Citations here) as there is often no guarantee that the right function was learnt during training.

This thesis involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter~ref{ch.4}). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks \cite{jaeger2001echo}, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

Formally, a function $T: U \to U$, where $U$ is some set is  a \emph{discrete-time dynamical system} and its iterates $\{u,Tu,T^2u,\ldots\}$, where $T^n$ denotes the $n$-fold composition of $T$ with itself, describe the evolution of an initial condition $u\in U$ (Note that we frequently drop the brackets and denote $T(u)$ by $Tu$ so as to simplify notations).  Also, we assume that $U$ is endowed with a metric $d$.

@@

\begin{Example} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $0<r\leq{4}$. It is of particular interest to us as it is classically used to illustrate chaos \cite{devaney2018introduction} for certain parameters. 
  In the graph below~\ref{fig:log_sdic} we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

  \begin{figure}[ht]
    \includegraphics[scale=0.74]{_logistic_sdic.eps}
        \centering
        \captionof{figure}{Figure generated for Logistic Map $x_{n+1}=rx_n(1-x_n)$ with $r=4$ to exhibit the presence of SDIC. Plotted are the values $x_n$ against time $n$ for the the first 50 step with initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.}
        \label{fig:log_sdic}
      \end{figure}

  % For an intuitive sense of the fact that the LM exhibits SDIC, consider again the definition and substitute specific values. Let $\delta=0.05$ and for purposes of illustration consider the value $u=0.4$. The green band in~\ref{fig:log_sdic} shades the region $[u-\delta, u+\delta]$ and we can see that regardless the value $v$  and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$
  % specific point, say $u=0.4$. Draw a green neighbourhood $[0.35,0.45]$ of diameters $\delta=0.02$ around the point. It is easy to see that every neighbourhood around $u=0.4$ will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta=0.05$ . 
\end{Example}

@@

\begin{Definition}\rm  
[\bf {Conjugacy \& Semi-Conjugacy}]\label{Dfn_Conjugate}\rm
   Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy with the underlying spaces. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$.
\end{Definition}

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly. M: Very often they are implicitly understood... so no worries, and I have edited.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold\footnote{A manifold is a special topological space that is locally homeomorphic at each point to an open set of $\mathbb{R}^d$} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

Hence, when the actual state $u$ of a system is seldom, if ever, fully known. the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: 
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently an orbit if discretisation  of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy (and so we could consider some noise to be involved).
\end{enumerate}

Takens embedding theorem does not take into account the noise, and for describing it, we construct a multidimensional observable using the method of stacking previous observations, i.e., we create delay-coordinate map defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
Takens' theorem, in essence, refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states evolving as $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

This requirement is not overly restrictive, since if a map is topologically transitive on a subspace it has to be surjective on it, since if it were to be not surjective there would be an open set $V$ that does not intersect with any future iterates of an open set  $U$ disjoint from it. \ednote{B: I remember you saying this in one of our sessions, but don't have a source for it. Do you have a specific reference? M: Follows from the transitivity definition. It is enough}

@@

The USP is equivalent called  the echo state property  \cite{Manju_ESP}, and systems that satisfy this property have other important and useful properties \cite{grigoryeva2018echo, grigoryeva2019differentiable, manjunath2020stability} for information processing.  
One of the first results obtained after defining the USP is that every solution will attract different initial conditions towards the component parts of the solution.
If $g$ has the USP, then any solution to $g$ is also a uniform attractor in nonautonomous dynamical systems literature~\cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this project, and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

@@

\begin{Definition}
  [\bf Causal Mapping]\label{Def_CausMap}
  A continuous, surjective map $H:\overleftarrow{U}\to\overleftarrow{X}_U$ such that \[H\circ\tilde{g}_v=\sigma_v \circ{H}\] holds for all $v \in U$ where $\tilde{g}_v$ maps $(\ldots, u_{-2}, u_{-1})$ to $(\ldots, u_{-2}, u_{-1}, g(v, u_{-1}))$ is called a causal mapping.
\end{Definition}

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The map $H^{-1}$ is continuous since $H$ is a continuous and surjective mapping of a compact space $\overleftarrow{U}$ to a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}.

@@

  It is imperative to take note of a subtlety here: we refer to the \emph{map} $H$ as a (causal) embedding, whereas when the driven \emph{system} $g$ (causally) embeds another \emph{system} $(U,T)$ we are concerned with $H_2$. \ednote{B: inserted line}

@@

%To achieve a causal embedding embedding not just a causal mapping, it is undesirable to choose a function $g$ that quenches the temporal structure in $u$ by contracting to such a degree that the ability to recover information from the original system is lost completely.
In the above example(\ref{ex_halfux}), the input's temporal variation could not be related to the reachable set as $X_U$ only consisted of a single element; little, if not no, information is encoded.
To obtain a suitably complex function $g$, it is thus desired that the reachable set of a driven system  be large enough to relate to the input. 
This is true even for embedding  the inverse-limit space of a dynamical system \cite{manjunath2021universal}.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.35]{_Nonlinearity_Fig_response.eps}
  \centering
\captionof{figure}{A coordinate of a simulated solution $(x_0,x_1,\ldots,x_{5000})$ of a RNN  plotted in red (with parameter $\alpha=0.99$ (where $g$ has the USP) and blue ($a=1$ and $\alpha=1.05$ where $g$ does not have the USP) while the matrices $A$ and $B$ are randomly generated, and $B$ has unit spectral radius. The input is shown in black. 
Reproduced from \cite{Manju_Nonlinearity}. }
\label{fig:memloss_conttime} 
\end{figure} 
\ednote{B: How can I explain the black line? I don't think this makes sense to the reader immediately. M: Oh it was the input in black that I have added.}

@@

In \cite{manjunath2021universal} it is illustrated in an empirical fashion in that the map $G_T$ describes dynamics which are less functionally complex than that of $T$ or of the map $\Phi_{2d,\theta}$. This is done by implementing a Recurrent Neural Network (RNN), which is later mentiond in Chapter~\ref{ch5}. \ednote{B: Should I add in that we'll be using Pearson coefficient? Not necessary. Please check all the cross citation. You need to explicitly mention that it is a Chapter or else only a number would appear}

@@

The reasons for taking this roundabout approach remain to be discussed in Section~\ref{subs_LearnGamma}, but a pat answer may immediately be given: When $\Gamma$ has been learnt, the system can be driven autonomously and then $G_T$ is known anyway. We formalise this with a theorem.

@@

Secondly, the function $\Gamma$ is known to be stable. Learning $\Gamma$ makes use of equations \ref{eqns_from_data} which in turn employs the driven system $g$ possessing the USP, and offers distinct advantages with regards to stability in the presence of perturbation. It is desirable that inputs that differ only slightly would have only ‘slight’ effects on the output of the system. Moreover, one would hope to prevent numerical errors originating from input and measurement noise (see~\cite[Th. 5]{manjunath2021universal}). 
As the data fed into the system is a function of both the input given and the parameters of the system itself (i.e. the value a, $\alpha$ in $\tanh$), we exploit both input-related and parameter-related stability. Both these notions are defined by way of the continuity of an encoding map of an input and design parameter respectively in \cite{manjunath2020stability}.

@@

Thirdly, learning a chaotic map on a set(here $Y_T$) with even a small measure of noise  means that the learnt system accepts arguments out of the set (Problems similar to Takens, cf. Section~\ref{sect_takenslimits}) \ednote{Please insert the right label}. Suppose iterates of  $(x_{n-1},x_{n})$ under the map $G_T$ is used to forecast two successive values of a solution of $g$ contained in $Y_T$, then these iterates would be contained in $Y_T$. However, in practice we are confronted with noise. Due to noise at some moment, an iterate could slip beyond $Y_T$ and into an 
expansion $Y_T^+$, for example a closed set that contains all the iterates of $(x_{n-1},x_{n})$ under the map $G_T^+$, where $G_T^+$ acts on $Y_T^+$.

Now since $Y_T$ is not guaranteed to be an attractor of $G_T^+$ this could lead to errors. To see this, consider Fig.~\ref{fig:YtGtFailure}.
Thus if one learns $G_T$ directly, we are in danger of replicating the problems arising for Takens.

\begin{figure}[ht]\includegraphics[scale=0.25]{_YTerrors.eps}
  \centering
  \captionof{figure}{Diagram of potential errors developing when $G_T$ is learnt. (Figure from \cite{Supp}}
  \label{fig:YtGtFailure}
\end{figure}

To circumvent this: if a driven system with a saturation function \footnote{a function $f(t)$ that is monotonic and approaches a constant as $|t|\to \infty$, for e.g., $\tanh(t)$ on $\mathbb{R}$} is used, we can ensure the existence of an attractor $A$ of the map that implements 
$(\pi_2,g) \circ (\Gamma,\pi_2)$ (printed in the red in ...\ednote{Please cite the commutativity diagram on the previous page}) on 
 which in practice turns out to be defined on $Y_T^+$ has an attractor $A$  containing $Y_T$.  \ednote{B: A little vague. Are we saying that an attractor of a map implementing $G_T$ is also the one that contains $Y_T^+$? M: Please see now.. there was typo.. it was A containing $Y_T$. I have expanded it now}
By restricting the dynamics to not spill beyond $A$, prevents large errors from occuring.  
More details on the global dissipativity of $g$, as it is termed here, may be found in~\cite{Supp}.

@@

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\ref{eqns_from_data}. To reach this point, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value and lies within the range $[-1,1]$ and required zero-padding(see paragraph below for further discussion), we continue by driving the system with external input through the function $g$.

@@

We make use of a Recurrent Neural Network (RNN) in our implementation to project the data on to a higher dimensional space, but we do not train the RNN in the familiar ways in the literature but as in \cite{manjunath2021universal}. A few words about the RNNs: RNNs are widely used in the literature for temporal processing of data as a type of deep learning technique. They take the form of equations considered to be discrete-time version of continuous dynamical systems modelling a neural network of neurons \cite{jaeger2001echo}.

We implement a feedforward network, (a network with no memory, unlike the RNN) required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library.

@@

The two model equations for learning from data~\eqref{eqns_from_data} allows us to determine what the state $x_{n+1}$ at time $n+1$ will be.

%We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.95\linewidth]{Graphs/_dp_success_4coords_traj.pdf} 
  \captionof{figure}{Predicted(red) and actual(blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads. From top to bottom are as follows: $x$-coordinate of topmost pendulum head, $y$-coordinate of topmost pendulum head, $x$-coordinate of lower pendulum head, $y$-coordinate of lower pendulum head. 
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 300 timesteps and stay close up until about 600 seconds.  } 
  \label{fig:dp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{_dp_success_4coords_hist.eps}
  \captionof{figure}{Densities of predicted and actual values of $x$- and $y$-coordinate of pendulum heads.}
  \label{fig:dp_success_density}
 \end{figure}
